Time Stretching of the GeV Emission of GRBs: Fermi LAT data vs geometrical model

### 2014-08-22 Paper/gammaRays.tex
Changed section names
Also increased the number of energy bins
Added "GRB " to GRB names

	In this paper we consider all the bursts observed by Fermi-LAT since 2008 August 4 to 2011 August 1, for which at least $10$ events with energies $\unit[1]{GeV}$ or higher were observed.
	There are $4$ of them: 080916C, 090510, 090902B, and 090926A.
	We study their light curves in two bands, $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$.

	The Kolmogorov-Smirnov test is used to check whether the light curves for these two bands are the same.
	No significant difference was found for 080916C and 090510.
	However, we observed with statistical significance of $3.3 \sigma$, that the higher energy light curve of 090926A is stretched with respect to the lower-energy one, and with statistical significance of $2.2 \sigma$, that the lower energy light curve of 090902B is stretched with respect to the higher-energy one.

@@

	Most of the bursts in the catalog, however, do not have enough events with over $\unit[1]{GeV}$ energies to do desired computations, so we choose only those bursts, from which at least $10$ photons were detected in the high energy band.
	It leaves $4$ of them: 080916C \cite{Tajima:2009az}, 090510 \cite{Ackermann:2010us}, 090902B \cite{Abdo:2009pg} and 090926A \cite{Bregeon:2011bu}.

@@

	Finally, we stretch the high energy CDF by different factors, and compare it to the low energy CDF using the KS-test.
	If we require $2\sigma$-significant probability to exclude a particular stretching factor, then the stretching factor of $1$ (which means no stretching) is allowed for GRBs 080916C and 090510.
	For the other two bursts the stretching factor of 1 is, however, excluded.
	For GRB 090902B it should be smaller than 1, so the low energy light curve is stretched with respect to the high energy one (see fig. \ref{fig:grb090902B}).
	And for GRB 090926A even more significant deviation from the stretching factor of 1 is observed in the opposite direction.
	The high energy light curve of GRB 090926A is stretched with respect to the low energy one by a factor of at least 1.99 (see fig. \ref{fig:grb090926A}).
	All observed stretching factors are summarized in table \ref{tab:observationResults}.

@@

	Finally, our assumption is able to explain the non-trivial stretching factor of both GRBs 090902B and 090926A.
	Detailed computations will be described in the following sections, but you can see the main idea on figure \ref{fig:modelOverview}.
	The computations confirm that this qualitative picture is indeed correct.
	You can see a sample light curves on figure \ref{fig:sampleLightCurves}, and predicted distribution of stretching factors on figure \ref{fig:kappaDistribution}.

@@

\subsection{Results}

@@

Out of these 4 bursts, two (080916C and 090510) have stretching factors compatible with $\kappa = 1$ within $2\sigma$ range.

@@

@@

The cost function is then computed by the following procedure:
\begin{enumerate}
	\item{
		Set $\gamma = 300$, $\omega_0 = \unit[1]{GeV}$, $z = 2.1062$ (which is the redshift of GRB 090926A)
	}
	\item{
		Set $\eta_0$ and $r_0$ such that the duration of the burst, and the total number of observed photons are compatible with GRB 090926A, $T_{0.99}\left(z,\chi;\unit[0.1]{GeV},\infty\right) = \unit[219.5]{sec}$ and $p_{\infty} \left(z,\chi;\unit[0.1]{GeV},\infty\right) = 179.989$
	}
	\item{
		If $k+\frac{\alpha}{2}+1 < 0$, or the energy of the burst $E\left(\unit[0.1]{GeV}\right) > \unit[6 \times 10^{53}]{GeV}$, the cost function equals to the penalization factor: $C\left(n, \theta_0, k, \alpha, \chi\right) = 400$.
	}
	\item{
		\label{item:sample}
		Compute the small sample of 10 bursts (we will need stretching factors and total photon counts) with the same fixed burst parameters $\left(\gamma, \eta_0, r_0, n, \omega_0, \theta_0, k, \alpha\right)$, and with $z$ and $\chi$ representatively distributed (as discussed in section \ref{sec:distribution}).
	}
	\item{
		Compute the cost due to the stretching factor:
		\begin{equation}
			C_\kappa = \frac{
				\log{\kappa\left(z,\chi;\unit[0.1]{GeV},\unit[1]{GeV},\infty\right)} - \log{\kappa_{090926A}}
			}{\Delta\log\kappa_{090926A}}
		\end{equation}
		Here $\log\kappa_{090926A} = \frac{1}{2}\left(\log\left(6.62\right) + \log\left(1.99\right)\right)$ and $\Delta\log\kappa_{090926A} = \frac{1}{2}\left(\log\left(6.62\right) - \log\left(1.99\right)\right)$.
	}
	\item{
		Compute the cost due to the minimal stretching factor from the sample:
		\begin{equation}
			C_{\kappa\text{min}} = \max\left(0, 
				\frac{
					\log{\kappa_\text{min}} - \log{\kappa_{090902B}}
				}{\Delta\log\kappa_{090902B}}\right)
		\end{equation}
		Here $\kappa_\text{min}$ is the minimal stretching factor from the sample computed in the step \ref{item:sample}, $\log\kappa_{090902B} = \frac{1}{2}\left(\log\left(0.89\right) + \log\left(0.36\right)\right)$ and $\Delta\log\kappa_{090902B} = \frac{1}{2}\left(\log\left(0.89\right) - \log\left(0.36\right)\right)$.
	}
	\item{
		Compute the cost due to the fraction of high and low energy photon counts:
		\begin{equation}
			C_f = \frac{
				\log{\frac{
					p_{\infty} \left(z,\chi;\unit[1]{GeV},\infty\right)
				}{
					p_{\infty} \left(z,\chi;\unit[0.1]{GeV},\unit[1]{GeV}\right)}
				} - \log f_{090926A}
			}{\log(10)}
		\end{equation}
		Here $f_{090926A} = 0.0606777$.
	}
	\item{
		Compute the cost due to the brightness of the burst compared to the median:
		\begin{equation}
			C_b = \frac{
				\log{\frac{
					p_{\infty} \left(z,\chi;\unit[0.1]{GeV},\infty\right)
				}{
					p_\text{med}
				}} - 0
			}{\log(10)}
		\end{equation}
		Here $p_\text{med}$ is the median number of observed photons among the sample computed in the step \ref{item:sample}.
	}
	\item{
		Finally, the value of the cost function is the sum of squares of the four:
		\begin{equation}
			C = C_\kappa^2 + C_{\kappa\text{min}}^2 + C_f^2 + C_b^2
		\end{equation}
	}
\end{enumerate}

@@

\subsection{Results}

@@

	So our model is able to explain the time stretching result from section \ref{sec:observations}.

>>>>

	In this paper we consider all the bursts observed by Fermi-LAT since 2008 August 4 to 2011 August 1, for which at least $10$ events with energies $\unit[1]{GeV}$ or higher were observed.
	There are $4$ of them: GRB 080916C, GRB 090510, GRB 090902B, and GRB 090926A.
	We study their light curves in two bands, $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$.

	The Kolmogorov-Smirnov test is used to check whether the light curves for these two bands are the same.
	No significant difference was found for GRB 080916C and GRB 090510.
	However, we observed with statistical significance of $3.3 \sigma$, that the higher energy light curve of GRB 090926A is stretched with respect to the lower-energy one, and with statistical significance of $2.2 \sigma$, that the lower energy light curve of GRB 090902B is stretched with respect to the higher-energy one.

@@

	Most of the bursts in the catalog, however, do not have enough events with over $\unit[1]{GeV}$ energies to do desired computations, so we choose only those bursts, from which at least $10$ photons were detected in the high energy band.
	It leaves $4$ of them: GRB 080916C \cite{Tajima:2009az}, GRB 090510 \cite{Ackermann:2010us}, GRB 090902B \cite{Abdo:2009pg} and GRB 090926A \cite{Bregeon:2011bu}.

@@

	Finally, we stretch the high energy CDF by different factors, and compare it to the low energy CDF using the KS-test.
	If we require $2\sigma$-significant probability to exclude a particular stretching factor, then the stretching factor of $1$ (which means no stretching) is allowed for GRB 080916C and GRB 090510.
	For the other two bursts the stretching factor of 1 is, however, excluded.
	For GRB 090902B it should be smaller than 1, so the low energy light curve is stretched with respect to the high energy one (see fig. \ref{fig:grb090902B}).
	And for GRB 090926A even more significant deviation from the stretching factor of 1 is observed in the opposite direction.
	The high energy light curve of GRB 090926A is stretched with respect to the low energy one by a factor of at least 1.99 (see fig. \ref{fig:grb090926A}).
	All observed stretching factors are summarized in table \ref{tab:observationResults}.

@@

	Finally, our assumption is able to explain the non-trivial stretching factor of both GRB 090902B and GRB 090926A.
	Detailed computations will be described in the following sections, but you can see the main idea on figure \ref{fig:modelOverview}.
	The computations confirm that this qualitative picture is indeed correct.
	You can see a sample light curves on figure \ref{fig:sampleLightCurves}, and predicted distribution of stretching factors on figure \ref{fig:kappaDistribution}.

@@

\subsection{Stretching factors}

@@

Out of these 4 bursts, two (GRB 080916C and GRB 090510) have stretching factors compatible with $\kappa = 1$ within $2\sigma$ range.

@@

\section{Results}

@@

The cost function is then computed by the following procedure:
\begin{enumerate}
	\item{
		Set $\gamma = 300$, $\omega_0 = \unit[1]{GeV}$, $z = 2.1062$ (which is the redshift of GRB 090926A)
	}
	\item{
		Set $\eta_0$ and $r_0$ such that the duration of the burst, and the total number of observed photons are compatible with GRB 090926A, $T_{0.99}\left(z,\chi;\unit[0.1]{GeV},\infty\right) = \unit[219.5]{sec}$ and $p_{\infty} \left(z,\chi;\unit[0.1]{GeV},\infty\right) = 179.989$
	}
	\item{
		If $k+\frac{\alpha}{2}+1 < 0$, or the energy of the burst $E\left(\unit[0.1]{GeV}\right) > \unit[6 \times 10^{53}]{GeV}$, the cost function equals to the penalization factor: $C\left(n, \theta_0, k, \alpha, \chi\right) = 400$.
	}
	\item{
		\label{item:sample}
		Compute the small sample of 10 bursts (we will need stretching factors and total photon counts) with the same fixed burst parameters $\left(\gamma, \eta_0, r_0, n, \omega_0, \theta_0, k, \alpha\right)$, and with $z$ and $\chi$ representatively distributed (as discussed in section \ref{sec:distribution}).
	}
	\item{
		Compute the cost due to the stretching factor:
		\begin{equation}
			C_\kappa = \frac{
				\log{\kappa\left(z,\chi;\unit[0.1]{GeV},\unit[1]{GeV},\infty\right)} - \log{\kappa_{\text{GRB090926A}}}
			}{\Delta\log\kappa_{\text{GRB090926A}}}
		\end{equation}
		Here $\log\kappa_{\text{GRB090926A}} = \frac{1}{2}\left(\log\left(6.62\right) + \log\left(1.99\right)\right)$ and $\Delta\log\kappa_{\text{GRB090926A}} = \frac{1}{2}\left(\log\left(6.62\right) - \log\left(1.99\right)\right)$.
	}
	\item{
		Compute the cost due to the minimal stretching factor from the sample:
		\begin{equation}
			C_{\kappa\text{min}} = \max\left(0, 
				\frac{
					\log{\kappa_\text{min}} - \log{\kappa_{\text{GRB090902B}}}
				}{\Delta\log\kappa_{\text{GRB090902B}}}\right)
		\end{equation}
		Here $\kappa_\text{min}$ is the minimal stretching factor from the sample computed in the step \ref{item:sample}, $\log\kappa_{\text{GRB090902B}} = \frac{1}{2}\left(\log\left(0.89\right) + \log\left(0.36\right)\right)$ and $\Delta\log\kappa_{\text{GRB090902B}} = \frac{1}{2}\left(\log\left(0.89\right) - \log\left(0.36\right)\right)$.
	}
	\item{
		Compute the cost due to the fraction of high and low energy photon counts:
		\begin{equation}
			C_f = \frac{
				\log{\frac{
					p_{\infty} \left(z,\chi;\unit[1]{GeV},\infty\right)
				}{
					p_{\infty} \left(z,\chi;\unit[0.1]{GeV},\unit[1]{GeV}\right)}
				} - \log f_{\text{GRB090926A}}
			}{\log(10)}
		\end{equation}
		Here $f_{\text{GRB090926A}} = 0.0606777$.
	}
	\item{
		Compute the cost due to the brightness of the burst compared to the median:
		\begin{equation}
			C_b = \frac{
				\log{\frac{
					p_{\infty} \left(z,\chi;\unit[0.1]{GeV},\infty\right)
				}{
					p_\text{med}
				}} - 0
			}{\log(10)}
		\end{equation}
		Here $p_\text{med}$ is the median number of observed photons among the sample computed in the step \ref{item:sample}.
	}
	\item{
		Finally, the value of the cost function is the sum of squares of the four:
		\begin{equation}
			C = C_\kappa^2 + C_{\kappa\text{min}}^2 + C_f^2 + C_b^2
		\end{equation}
	}
\end{enumerate}

@@

\subsection{Results of the tests}

@@

	So our model is able to explain the time stretching effect from section \ref{sec:observations}.
